Article Title

### 2014-10-29 manuscript.tex
added Ising model

Given any covariance matrix estimate $C$, the corresponding correlation matrix $R$ is calculated by normalizing the rows and columns of $C$ by the square roots of its diagonal elements to produce unit entries on the diagonal:
\begin{equation}\label{eq:precision}
    R = \left(I\circ C\right)^{-\frac 1 2} C \left(I\circ C\right)^{-\frac 1 2},
\end{equation}
where $\circ$ denotes the entrywise matrix product (Hadamard product) and $I$ is the $p\times p$ identity matrix.

@@

Pairwise partial correlations are closely related to the elements of the \emph{precision matrix}, \emph{i.e.}\;the inverse of the covariance matrix \cite{Dempster:1972,Whittaker:1990}. Zero elements in the precision matrix signify zero partial correlation between the two variables. Given the covariance estimate $C$, the matrix of partial correlations $P$ is computed by normalizing the rows and columns of the \emph{precision matrix} $C^{-1}$ to produce negative unit entries on the diagonal:
\begin{equation}\label{eq:partial}
    P = -\left(I\circ C^{-1}\right)^{-\frac 1 2} C^{-1} \left(I\circ C^{-1}\right)^{-\frac 1 2}
\end{equation}

@@

Estimator $C_{\sf sparse+latent}$ is obtained by approximating the sample covariance matrix by a matrix whose inverse is the sum of a sparse component and a low-rank component:
\begin{equation}\label{eq:c-sl}
C_{\sf sparse+latent} = (S - L)^{-1},
\end{equation}
where, as above, $S$ is a sparse matrix and $L$ is a low-rank matrix. The estimator has two hyperparameters: one to regulate the sparsity of $S$ and the other to regulate the rank of $L$. See Methods for a more detailed explanation. The structure imposed by $C_{\sf sparse+latent}$ favors conditions in which the activity of neurons is determined by linear effects between some observed pairs of neurons and linear effects from several latent units (Fig.~\ref{fig:1} Row 1, D) \cite{Chandrasekaran:2010,Ma:2013}.

@@

\subsection*{Simulation}
We next demonstrate how the most efficient among different regularized estimators can reveal the structure of correlations.
We constructed four families of $50\times 50$ covariance matrices, each with structure that matched one of the four regularized estimators (Fig.~\ref{fig:1} Row 2, A--D and Methods).  We used these covariance matrices as the ground truth in multivariate normal distributions with zero means and drew samples of various sizes. Even with relatively large sample sizes (\emph{e.g.}\;$n=500$), the sample correlation matrices were contaminated by sampling noise and their underlying structures were difficult to discern (Fig.~\ref{fig:1} Row 3).

@@

Cross-validation loss accurately reproduced the relationships between the excess losses of the estimators (Fig.~\ref{fig:1} Row 6), confirming that cross-validation can evaluate true loss.

@@

To test this hypothesis, we computed the relative cross-validation loss of estimators  $C_{\sf sample}$, $C_{\sf diag}$, $C_{\sf factor}$, and $C_{\sf sparse}$ with respect to $C_{\sf sparse+latent}$ in $n=27$ imaged sites in 14 mice.  The hyperparameters of each estimator were optimized by nested cross-validation (See Fig.~S1 and  Methods). Indeed, the sparse+latent estimator outperformed the other estimators (Fig.~\ref{fig:3}). The respective median differences of the cross-validation loss were 0.039, 0.0016, 0.0029, and 0.0059 nats/cell/bin, significantly greater than zero ($p<0.01$ in each comparison, $n=27$ sites in 14 mice, Wilcoxon signed rank test).

The same evaluation based on the quadratic loss function,
\begin{equation}\label{eq:quad}
\loss{C,C_{\sf sample}^\prime}=\frac 1 {p^2}\Tr(C^{-1}C_{\sf sample}^\prime-I)^2,
\end{equation}
instead of the normal loss function reproduced the same relationship between the estimators (Fig.~S2). This suggests that the results of this study are robust to the choice of the loss function and do not depend on the assumption of gaussianity.

@@

We evaluated the covariance matrix estimators a loss function based on multivariate normal log likelihood.  However, this does not limit the applicability of its conclusions to normal distributions. Indeed our major findings could be reproduced using other loss functions (compare Fig.~\ref{fig:3} and Fig.~S2).  Other probabilistic models, fitted to the same data, could also serve as estimators of the covariance matrix.  If a different model yields better estimation of the covariance matrix than the estimator proposed here, we believe that its structure should deserve consideration as the better representation of the functional connectivity.

@@

Since each of the regularized estimators had one or two hyperparameters, we used \emph{nested cross-validation}:  The outer loop evaluated the performance of the estimators with the hyperparameter values optimized by cross-validation within the inner loop.  Hyperparameters were optimized by a two-phase search algorithm: random search to find a good starting point for the subsequent pattern search to find the global minimum.  The inner cross-validation loop subdivided the training dataset from the outer loop to perform 10-fold cross-validation in order to evaluate each choice of the hyperparameter values.  Thus the size of the training dataset within the inner loop comprised 81\% of the entire recording. Fig.~S1 illustrates the dependence of the validation loss on the hyperparameters of the $C_{\sf sparse+latent}$ estimator for the example site shown in Figures \ref{fig:2} and \ref{fig:4} and the optimal value found by the pattern search algorithm.

@@

Estimator $C_{\sf diag}$ (Eq.~\ref{eq:c-diag})  used two hyperparameters: the covariance shrinkage intensity $\lambda \in [0,1]$ and variance shrinkage intensity $\alpha \in [0,1]$.  The variances (the diagonal of $C_{\sf sample}$) were shrunk linearly toward their mean value $\frac 1 p \Tr(C_{\sf sample})$:
\begin{equation}
D = (1-\alpha)C_{\sf sample}\circ I + \alpha \frac 1 p \Tr(C_{\sf sample}) I
\end{equation}
The $C_{\sf diag}$ estimate was then obtained by shrinking $C_{\sf sample}$ toward $D$ according to Eq.~\ref{eq:c-diag}.

@@

The partial correlation matrix (Eq.~\ref{eq:partial}) computed from $C_{\sf sparse+latent}$ includes interactions between the visible and latent units and was used in Fig.~\ref{fig:4} C and D and Fig.~\ref{fig:5} C, and Fig.~\ref{fig:6} A--C).  The partial correlation matrix computed from $S$ alone expresses strengths of pairwise interactions
\begin{equation}
P_{\sf sparse} = -(S\circ I)^{-\frac 1 2} S  (S\circ I)^{-\frac 1 2}
\end{equation}
and were used in Fig.~\ref{fig:4} F, G, H.

@@

The common correlation matrix $R_{\sf sample}$ is estimated by averaging the condition-specific correlations:
\begin{equation}
    R_{\sf sample}  
    = \frac 1 n \sum\limits_c n_c \left(V_{c,\sf sample}^{-\frac 1 2}C_{c,\sf sample}V_{c,\sf sample}^{-\frac 1 2}\right)
    = \frac 1 n \sum\limits_c \sum\limits_{t \in T_c} z(t)z(t)^\T,
\end{equation}
where $n=\sum\limits_c n_c$ and $V_{c,\sf sample} = C_{c,\sf sample} \circ I$ is the diagonal matrix containing the sample variances. Then $R_{\sf sample}$ is simply the covariance matrix of the $z$-score signal $z(t) = V_{c,\sf sample}^{-\frac 1 2} \left(x(t) - \bar x_c\right)$ of the training sample.

@@

%\section*{References}
% Either type in your references using
% \begin{thebibliography}{}
% \bibitem{}
% Text
% \end{thebibliography}
%
% OR
%
% Compile your BiBTeX database using our plos2009.bst
% style file and paste the contents of your .bbl file
% here.
% 
\bibliography{references.bib}

@@

\newpage
\section*{Supporting Information Legends}
%
% Please enter your Supporting Information captions below in the following format:
%\item{\bf Figure SX. Enter mandatory title here.} Enter optional descriptive information here.
% 
\begin{description}
\item {\bf Figure S1. Optimization of hyperparameters of the $C_{\sf sparse+latent}$ estimator.} {\bf A.} Validation loss (Eq.~\ref{eq:full-loss}) for the example site in Fig.~\ref{fig:2} and \ref{fig:4} as a function of the hyperparameters $\alpha$ and $\beta$ of the $C_{\sf sparse+latent}$ estimator (Eq.~\ref{eq:c-sl} and Eq.~\ref{eq:ma}). In all panels, the red cross marks the optimal value found by the pattern search algorithm described in Methods.
{\bf B.} The connectivity ($1-\mbox{sparsity}$) of the sparse component $S$ as a function of $\alpha$ and $\beta$ for the example site.
{\bf C.} The number of latent units, \emph{i.e.}~the rank of the low-rank component $L$, as a function of hyperparameters $\alpha$ and $\beta$.
{\bf D.} The loss function as a function of the connectivity and the number of latent units.

\item {\bf Figure S2. Performance of estimator $C_{\sf sparse+latent}$ with respect to \emph{quadratic loss} (Eq.~\ref{eq:quad}) relative to the other estimators: $C_{\sf sample}$, $C_{\sf diag}$, $C_{\sf factor}$, and $C_{\sf sparse}$.} All comparisons showed significant advantage of estimator $C_{\sf sparse+latent}$ ($p<10^{-3}$ in each comparison, Wilcoxon signed rank test, $n=27$ sites in 14 mice). 
\end{description}

>>>>

Given any covariance matrix estimate $C$, the corresponding correlation matrix $R$ is calculated by normalizing the rows and columns of $C$ by the square roots of its diagonal elements to produce unit entries on the diagonal:
\begin{equation}\label{eq:precision}
    R = (\mbox{diag}(C))^{-\frac 1 2} C (\mbox{diag}(C))^{-\frac 1 2},
\end{equation}
where $\mbox{diag}(C)$ denotes the diagonal matrix with the diagonal elements from $C$.

@@

Pairwise partial correlations are closely related to the elements of the \emph{precision matrix}, \emph{i.e.}\;the inverse of the covariance matrix \cite{Dempster:1972,Whittaker:1990}. Zero elements in the precision matrix signify zero partial correlation between the two variables. Given the covariance estimate $C$, the matrix of partial correlations $P$ is computed by normalizing the rows and columns of the \emph{precision matrix} $C^{-1}$ to produce negative unit entries on the diagonal:
\begin{equation}\label{eq:partial}
    P = -\left(\mbox{diag}(C^{-1})\right)^{-\frac 1 2} C^{-1} \left(\mbox{diag}(C^{-1})\right)^{-\frac 1 2}
\end{equation}

@@

Estimator $C_{\sf sparse+latent}$ is obtained by approximating the sample covariance matrix by a matrix whose inverse is the sum of a sparse component and a low-rank component:
\begin{equation}\label{eq:c-sl}
C_{\sf sparse+latent} = (S - L)^{-1},
\end{equation}
where $S$ is a sparse matrix and $L$ is a low-rank matrix. The estimator has two hyperparameters: one to regulate the sparsity of $S$ and the other to regulate the rank of $L$. See Methods for a more detailed explanation. The structure imposed by $C_{\sf sparse+latent}$ favors conditions in which the activity of neurons is determined by linear effects between some observed pairs of neurons and linear effects from several latent units (Fig.~\ref{fig:1} Row 1, D) \cite{Chandrasekaran:2010,Ma:2013}.

@@

\subsection*{Simulation}
We next demonstrate how the most efficient among different regularized estimators can reveal the structure of correlations.
We constructed four families of $50\times 50$ covariance matrices, each with structure that matched one of the four regularized estimators (Fig.~\ref{fig:1} Row 2, A--D and Methods).  We used these covariance matrices as the ground truth in multivariate normal (MVN) distributions with zero means and drew samples of various sizes. 
The sample correlation matrices from finite samples (\emph{e.g.}\ $n=500$ in Fig.~\ref{fig:1} Row 3) were contaminated with sampling noise and their underlying structures were difficult to discern.

@@

Cross-validation loss accurately reproduced the ranking of estimators (Fig.~\ref{fig:1} Row 6), confirming that cross-validation can evaluate the true loss of the estimators.

To demonstrate that estimator rankings did not depend on the assumption of gaussianity, we repeated the cross-validated evaluation with a non-Gaussian data generating model, non-Gaussian loss function, or both:

First, we replaced the data-generating model with the pairwise Ising model \cite{Hertz:2011}, which has equivalent mathematical form to the MVN model,
\begin{equation}
    x\sim \frac 1 {Z(S,h)} \exp\left(-\frac 1 2 x^\T S x + h^\T x \right),
\end{equation}
but defined on the binary domain $x\in\{-1,+1\}^p$ rather than the continuous domain of the MVN model. The partition function $Z(S,h)$ normalizes the distribution on the models' respective domains. 
For the MVN, the matrix $S$ is the precision matrix or the inverse of the covariance matrix; and the mean values are $\mu=-S^{-1}h$.  In the Ising model, $-S$ is the matrix of pairwise interactions and $h$ is the vector of intrinsic activity drives of individual cells but $S$ and $h$ do not have a closed-form relationship to the mean activity and the covariance matrix of the distribution. 
However, both distributions have the same structure of pairwise conditional dependencies: zeros in the matrix $S$ indicate conditional independence between the corresponding pair of neurons.

First, replacing the normal loss function with the quadratic loss function, 
\begin{equation}\label{eq:quad}
\loss{C,C_{\sf sample}^\prime}=\frac 1 {p^2}\Tr(C^{-1}C_{\sf sample}^\prime-I)^2
\end{equation}
did not affect estimator rankings (Fig.~S1 A--D).

@@

To test this hypothesis, we computed the relative cross-validation loss of estimators  $C_{\sf sample}$, $C_{\sf diag}$, $C_{\sf factor}$, and $C_{\sf sparse}$ with respect to $C_{\sf sparse+latent}$ in $n=27$ imaged sites in 14 mice.  The hyperparameters of each estimator were optimized by nested cross-validation (See Fig.~S2 and  Methods). Indeed, the sparse+latent estimator outperformed the other estimators (Fig.~\ref{fig:3}). The respective median differences of the cross-validation loss were 0.039, 0.0016, 0.0029, and 0.0059 nats/cell/bin, significantly greater than zero ($p<0.01$ in each comparison, $n=27$ sites in 14 mice, Wilcoxon signed rank test).

The same evaluation based on the quadratic loss function,
instead of the normal loss function reproduced the same relationship between the estimators (Fig.~S3).

This suggests that the results of this study are robust to the choice of the loss function and do not depend on the assumption of gaussianity.

@@

We evaluated the covariance matrix estimators a loss function based on multivariate normal log likelihood.  However, this does not limit the applicability of its conclusions to normal distributions. Indeed our major findings could be reproduced using other loss functions (compare Fig.~\ref{fig:3} and Fig.~S3).  Other probabilistic models, fitted to the same data, could also serve as estimators of the covariance matrix.  If a different model yields better estimation of the covariance matrix than the estimator proposed here, we believe that its structure should deserve consideration as the better representation of the functional connectivity.

@@

Since each of the regularized estimators had one or two hyperparameters, we used \emph{nested cross-validation}:  The outer loop evaluated the performance of the estimators with the hyperparameter values optimized by cross-validation within the inner loop.  Hyperparameters were optimized by a two-phase search algorithm: random search to find a good starting point for the subsequent pattern search to find the global minimum.  The inner cross-validation loop subdivided the training dataset from the outer loop to perform 10-fold cross-validation in order to evaluate each choice of the hyperparameter values.  Thus the size of the training dataset within the inner loop comprised 81\% of the entire recording. Fig.~S2 illustrates the dependence of the validation loss on the hyperparameters of the $C_{\sf sparse+latent}$ estimator for the example site shown in Figures \ref{fig:2} and \ref{fig:4} and the optimal value found by the pattern search algorithm.

@@

Estimator $C_{\sf diag}$ (Eq.~\ref{eq:c-diag})  used two hyperparameters: the covariance shrinkage intensity $\lambda \in [0,1]$ and variance shrinkage intensity $\alpha \in [0,1]$.  The variances (the diagonal of $C_{\sf sample}$) were shrunk linearly toward their mean value $\frac 1 p \Tr(C_{\sf sample})$:
\begin{equation}
D = (1-\alpha)\mbox{diag}(C_{\sf sample}) + \alpha \frac 1 p \Tr(C_{\sf sample}) I
\end{equation}
The $C_{\sf diag}$ estimate was then obtained by shrinking $C_{\sf sample}$ toward $D$ according to Eq.~\ref{eq:c-diag}.

@@

The partial correlation matrix (Eq.~\ref{eq:partial}) computed from $C_{\sf sparse+latent}$ includes interactions between the visible and latent units and was used in Fig.~\ref{fig:4} C and D and Fig.~\ref{fig:5} C, and Fig.~\ref{fig:6} A--C).  The partial correlation matrix computed from $S$ alone expresses strengths of pairwise interactions
\begin{equation}
P_{\sf sparse} = -(\mbox{diag}(S))^{-\frac 1 2} S  (\mbox{diag}(S))^{-\frac 1 2}
\end{equation}
and were used in Fig.~\ref{fig:4} F, G, H.

@@

The common correlation matrix $R_{\sf sample}$ is estimated by averaging the condition-specific correlations:
\begin{equation}
    R_{\sf sample}  
    = \frac 1 n \sum\limits_c n_c \left(V_{c,\sf sample}^{-\frac 1 2}C_{c,\sf sample}V_{c,\sf sample}^{-\frac 1 2}\right)
    = \frac 1 n \sum\limits_c \sum\limits_{t \in T_c} z(t)z(t)^\T,
\end{equation}
where $n=\sum\limits_c n_c$ and $V_{c,\sf sample} = \mbox{diag}(C_{c,\sf sample})$ is the diagonal matrix containing the sample variances. Then $R_{\sf sample}$ is simply the covariance matrix of the $z$-score signal $z(t) = V_{c,\sf sample}^{-\frac 1 2} \left(x(t) - \bar x_c\right)$ of the training sample.

@@

%\section*{References}
% Either type in your references using
% \begin{thebibliography}{}
% \bibitem{}
% Text
% \end{thebibliography}
%
% OR
%
% Compile your BiBTeX database using our plos2009.bst
% style file and paste the contents of your .bbl file
% here.
% 
\bibliography{references}

@@

\newpage
\section*{Supporting Information Legends}
%
% Please enter your Supporting Information captions below in the following format:
%\item{\bf Figure SX. Enter mandatory title here.} Enter optional descriptive information here.
% 
\begin{description}
\item {\bf Figure S2. Optimization of hyperparameters of the $C_{\sf sparse+latent}$ estimator.} {\bf A.} Validation loss (Eq.~\ref{eq:full-loss}) for the example site in Fig.~\ref{fig:2} and \ref{fig:4} as a function of the hyperparameters $\alpha$ and $\beta$ of the $C_{\sf sparse+latent}$ estimator (Eq.~\ref{eq:c-sl} and Eq.~\ref{eq:ma}). In all panels, the red cross marks the optimal value found by the pattern search algorithm described in Methods.
{\bf B.} The connectivity ($1-\mbox{sparsity}$) of the sparse component $S$ as a function of $\alpha$ and $\beta$ for the example site.
{\bf C.} The number of latent units, \emph{i.e.}~the rank of the low-rank component $L$, as a function of hyperparameters $\alpha$ and $\beta$.
{\bf D.} The loss function as a function of the connectivity and the number of latent units.

\item {\bf Figure S3. Performance of estimator $C_{\sf sparse+latent}$ with respect to \emph{quadratic loss} (Eq.~\ref{eq:quad}) relative to the other estimators: $C_{\sf sample}$, $C_{\sf diag}$, $C_{\sf factor}$, and $C_{\sf sparse}$.} All comparisons showed significant advantage of estimator $C_{\sf sparse+latent}$ ($p<10^{-3}$ in each comparison, Wilcoxon signed rank test, $n=27$ sites in 14 mice). 
\end{description}
